KIN: A method to infer relatedness from low-coverage ancient DNA

### 2022-04-07 paper_overleaf/main.tex
Updates from Overleaf

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig3:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  and for ascertained data. One exception are  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias.. For simulations with ROH, we find the overall classification accuracy is only slightly reduced. Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN.

@@

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd} Degree$, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

@@

Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child. We also find that both  KIN and READ classify Chagyrskaya 06 and Chagyrskaya 14 as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).

We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d. $>$1) and KIN results ($\Delta LL>$1) match (fig. S\ref{figS6:Chagyrskaya_ibd}).  We find that lcMLkin does create four different clusters for different relatedness cases. However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.  Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at $r \approx 0.6, k_0 \approx 0.33$.

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree. Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2). However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Delta LL>1$. When we investigated some of these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig. S\ref{figS8:eg1}). Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs). Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).  
We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3). There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLkin differ, and again the heterozygosities along the genome match the inference of KIN (fig. S\ref{figS9:eg2}).

@@

In our simulations, we find that adding long ROH ($~17\%$) to simulations improves the performance of  both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias  (see Fig. \ref{fig3:Comparison_READ_KIN}).

@@

While we have focused on the application of KIN on ancient human samples, but the model is not tied to this system. In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. Thus, assuming we know the recombination rate, and hence can estimate the transition matrix \ref{method}, KIN can be widely applied to any diploid species and is easy to automatise and apply to large datasets.

@@

\paragraph{Transition matrices}
KIN requires a transition matrix for each relatedness case. which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Figure S \ref{figS10:pedigree}). For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:

For grandparent-grandchild, 
 $$Q = \left[\begin{array}
{rrr}
-r & r \\
r & -r 
\end{array}\right]$$

@@

*Remember to use updated lcMLkin in the plot.
\begin{figure}[h!]
    \centering
    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIN. Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
    The relatedness shown with different colors is estimated with both READ and KIN.}
    \label{figS6:Chagyrskaya_ibd}
\end{figure}

>>>>

\paragraph{Comparing KIN-HMM with READ}
In Figure Fig. \ref{fig3:Comparison_READ_KIN}, we compare the classification accuracy of KIN (cutoff: 1 log-likelihood unit) to that of READ (cutoff: 1 standard deviation)  under four different simulation scenarios. We show that both methods have similar performance for low-coverage shotgun data (``control''-case),  and for ascertained data. One exception is  $2^{nd}$-degree relatives, where KIN has higher power, especially at low coverages. We also find that READ is strongly impacted by contamination, but that the correction implemented in KIN is sufficient to remove this bias. For simulations with ROH, we find the overall classification accuracy is only slightly reduced. Since ROH makes individuals more similar, we see a bias towards more false positives, for first and second-degree relationships, which is slightly higher for READ than for KIN.

@@

\paragraph{Evaluating IBD prediction}
KIN-HMM co-estimates IBD states and relatedness, and hence it has a higher power to classify $2^{nd}$ Degree relatives, and can distinguish between Siblings and Parent-Child. We investigate our limit to predict IBD states along the genome by calculating the number of genomic windows where we correctly  predict IBD state for a pair of individuals. We calculate mean and standard deviation for this accuracy over 60 simulations of a pedigree (see section \ref{simulat}). Fig. \ref{fig4:IBDstate_accuracy} shows that we can predict IBD states for Unrelated, $3^{rd}$ Degree, $2^{nd}$ Degree, and Sibling pairs very well at high coverages, and our accuracy decreases with downsampling. On the other hand, our accuracy measure for Identical individuals and Parent-Child remains very high, due to robust relatedness estimates for these cases even at 0.03 x.

@@

Our estimates are consistent with those obtained using READ, except that READ is unable to detect 3rd-degree relationships, and cannot distinguish between siblings and parent-child. We also find that both  KIN and READ classify Chagyrskaya06 and Chagyrskaya14 as parent-child with low confidence, but from the morphology they likely stem from the same individual. We believe that this due to non-human contamination present in these libraries ($1\%$ and $2.9\%$ respectively).

We compared our IBD estimates to lcMLkin for all pairs for whom READ results (with s.d. $>$1) and KIN results ($\Delta LL>$1) match (fig. S\ref{figS6:Chagyrskaya_ibd}).  We find that lcMLkin does create four different clusters for different relatedness cases. However, the location of these clusters strongly deviate from the expected values, likely due to the low coverage and ROH.  Fossils from the same individual, for example, are expected to be at $k_0 = 0$ and $r = 1$, but are at k_0 \approx 0.33$, $r \approx 0.6.

@@

We further applied KIN to a genome-wide dataset of 118 ancient individuals from the Lech valley \cite{mittnik_kinship-based_2019}. We compared our relatedness estimates to those obtained from the authors using READ and lcMLkin, and found that in most cases the three methods agree. Out of 6903 pairwise comparisons, 346 pairs show different relatedness for KIN and READ (table S2). However, most of these are at very low confidence, and there are only 31 pairs where KIN has $\Delta LL>1$. When we investigated these comparisons in more detail we found that in many cases, the proportion of differences along the genome align with the prediction by KIN (fig. S\ref{figS8:eg1}). Only 20 pairs in this subset of 346 pairs had sufficient data for analysis using lcMLkin (more than 10,000 overlapping SNPs). Among these 20 pairs, we found that lcMLkin matches KIN's estimate in all cases except for pair POST131-POST28 (READ predicts unrelated individuals, lcMLkin predicts 3rd-5th degree, KIN predicts second degree) and pair AITI72-AITI77A (READ and lcMLkin predict unrelated, while KIN estimates 3rd degree with LL=1.7).  
We further found 43 pairs for which READ estimates First-Degree, and compared the estimates of lcMLkin and KIN for these cases (Table S3). There were a total of 20 pairs where lcMLkin had sufficient data, and 26 pairs where $\Delta LL>1$ for KIN. Among the 19 overlapping pairs, we found four pairs where KIN and lcMLkin differ, and again the heterozygosities along the genome match the inference of KIN (fig. S\ref{figS9:eg2}).

@@

In our simulations, we find that adding long ROH ($~17\%$) to simulations slightly improves the performance of  both KIN and READ, particularly at lower coverages. This is likely because ROH actually reduces the variance in differences, and so it becomes easier to estimate the location of IBD tracts. For READ, the presence of ROH causes a bias towards inferring closer relatedness cases, but the model we use in KIN reduces this bias (see Fig. \ref{fig3:Comparison_READ_KIN}).

@@

While we have focused on the application of KIN on ancient human samples, but the model is not tied to this system. In addition, the output of KIN is a table which shows for each pair, the most likely model, and the second best guess, along with a confidence level represented by the log likelihood ratio. Thus, assuming we know the recombination rate, and hence can estimate the transition matrix (see section \ref{method}), KIN can be widely applied to any diploid species and is easy to automatise and apply to large datasets.

@@

\paragraph{Transition matrices}
KIN requires a transition matrix for each relatedness case. which we estimate  by counting the transition between IBD states for all pairs of individuals at that relatedness level in a training set of 1000 simulations from our pedigree (Fig. S \ref{figS10:pedigree}). For two cases, siblings and grandparent-grandchild, it is possible to write down the theoretical expectation of the transition matrix:

For grandparent-grandchild, rate matrix is
 $$Q = \left[\begin{array}
{rrr}
-r & r \\
r & -r 
\end{array}\right]$$

@@

\begin{figure}[h!]
    \centering
    \includegraphics[width=18cm]{supplementary_info/plots/lcPlot.png}
    \caption{Comparison of IBD states estimated for Chagyrskaya specimens using (A) lcMLkin and (B) KIN. Each plot shows relatedness coefficient (calculated as $k_2+k_1/2$) on y-axis plotted against $k_0$ (proportion of genome with no chromosomes shared).
    The relatedness shown with different colors is estimated with both READ and KIN.}
    \label{figS6:Chagyrskaya_ibd}
\end{figure}
